---
title: "R Notebook"
output: html_notebook
---

# set-up

```{r}
rm(list = ls())
gc()
.rs.restartR()
```

# packages

```{r}
suppressPackageStartupMessages({library(tidyverse)
                               library(data.table)
                               library(caper)
                               library(ape)
                               library(ggpubr)})

today <- Sys.Date()

# global plotting theme
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white")
    )
)
```

# compile jules' lists

```{r}
# load 
pt1 <- fread("../../data/fastq_master_mapped_pt1.csv")
pt2 <- fread("../../data/fastq_species_master_mapped_part2_filtered.csv")
all_sam <- rbind(pt1, pt2)

# basic stats
species_with_many_samples <- all_sam %>%
  group_by(Species) %>%
  filter(n() > 1)

# list of species where runs collapsed to biosamples
collapsed_sam_species <- unique(all_sam[grepl("SAMN", all_sam$SRR), "Species"])

# runs for species without collapsed samples
non_collapsed_sam <- all_sam[!(all_sam$Species %in% collapsed_sam_species$Species),]
```

## output biosamples to be processed with entrez

```{r}
# output SAMN
write.table(all_sam[grepl("SAMN|SAME|SAMD", all_sam$SRR), "SRR"], "../../resources/biosample_ids.txt", quote = F, row.names = F, col.names = F)
```

## now grab run ids associated with biosamples

## get SRR associated with SAMN 

remove any SRR that are not matching jules' filters

awk -F, 'NR==1 || ($13 == "WGS" && $15 == "GENOMIC" && $16 == "PAIRED" && $19 == "ILLUMINA")' SraRunInfo.csv > filtered_SraRunInfo.csv

```{r}
runinfo <- fread("/global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources/runinfo.csv")

runinfo_filt <- runinfo %>% filter(LibraryStrategy == "WGS" & LibrarySource == "GENOMIC" & LibraryLayout == "PAIRED" & Platform == "ILLUMINA")

all_sam_run_only <- merge(all_sam, runinfo_filt[,c("Run", "BioSample")], by.x = "SRR", by.y = "BioSample", all.x = T)

# if run starts with SAMN, replace with real Run number
samn_bool <- grepl("SAMN|SAME|SAMD", all_sam_run_only$SRR)
all_sam_run_only$SRR[samn_bool] <- all_sam_run_only$Run[samn_bool]

# expect TRUE
if(!all(all_sam_run_only$SRR == all_sam_run_only$Run, na.rm = T)){
  stop("SRR does not match Run")
}

```

## now output the runs and grab all of their biosample ids

My hope is to standardize the info, so it's intuitive, rather than having my grouping variable sometimes be the run id and sometimes be the biosample id

```{r}
write.table(all_sam_run_only[,1], "../../resources/runs_srr_only.txt", quote = F, row.names = F, col.names = F)
```

## now grab all run info for every run

```{bash, engine.opts='-l'}

cd /global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources
pwd

ENV_BIN="/global/home/users/milesroberts/.local/share/mamba/envs/entrez/bin"

readarray -t runs < "runs_srr_only.txt"

# Clear out the old file if it exists so you don't keep appending duplicates
> runinfo_all_runs.csv

for run in "${runs[@]}"; do
    echo "$run"
    $ENV_BIN/esearch -db sra -query $run | $ENV_BIN/efetch -format runinfo >> runinfo_all_runs.csv
    sleep 1
done
```

## Finally, organize runinfo into one table with three columns: run, biosample, species

```{r}
runinfo <- read.csv("/global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources/runinfo_all_runs_unique.csv", fill = T, header = F)

# run_list <- read.csv("/global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources/runs_srr_only.txt", fill = T, header = F)
# 
# sum(run_list$V1 %in% runinfo$V1)

runinfo_sorted <- runinfo %>% 
  #filter(!grepl("metagenome|uncultured|unidentified|human|synthetic|prokaryotic|mixed libraries|Mus musculus|Homo sapiens|Saccharomyces cerevisiae|Saccharomonospora azurea|Caenorhabditis elegans|Bos taurus|Gallus gallus|Monodelphis domestica|Drosophila", V29)) %>%
  #filter(V29 %in% (all_sam$Species)) %>% 
  filter(V13 == "WGS" & V15 == "GENOMIC" & V16 == "PAIRED" & V19 == "ILLUMINA") %>%
  dplyr::select(V1, V26, V29) %>% 
  arrange(V26, V1, V29) 

runinfo_sorted <- runinfo_sorted[-1:-5,]

# replace space with _
runinfo_sorted$V29 <- gsub(" ", "_", runinfo_sorted$V29)
names(runinfo_sorted) = c("Run", "BioSample", "Species")

# remove species with only one run
# cannot calculate k-mer distances for such species
# sort species by number of runs (analyze species with less data first)
runinfo_filtered <- runinfo_sorted %>%
  group_by(Species) %>%
  #filter(n_distinct(BioSample) > 1) %>%
  ungroup() %>%
  droplevels() %>%
  filter(Species %in% (all_sam$Species)) %>% 
  arrange(Species)

write.table(runinfo_filtered, "../../config/samples_gte1Biosample.tsv", quote = F, row.names = F, col.names = T, sep = "\t")

runinfo_filtered <- runinfo_filtered %>%
  filter(n_distinct(BioSample) > 1)

# summary statistics
cat("Number of Species:", n_distinct(runinfo_filtered$Species), "\n")
cat("Number of BioSamples:", n_distinct(runinfo_filtered$BioSample), "\n")
cat("Number of Runs:", n_distinct(runinfo_filtered$Run), "\n")

# check for species not in original Jules list
# seems like my list excludes some species that are cultivars
runinfo_filtered$Species[!(runinfo_filtered$Species %in% species_with_many_samples$Species)]

unique(species_with_many_samples$Species[!(species_with_many_samples$Species %in% runinfo_filtered$Species)])

# save list of species
write.table(runinfo_filtered, "../../config/samples_all.tsv", quote = F, row.names = F, col.names = T, sep = "\t")

# partition species into groups with a small, medium, and large amount of data
# will process species with very little data first
runinfo_partitioned <- runinfo_filtered %>%
  group_by(Species) %>%
  mutate(num_runs = n(),
         batches = case_when(
    num_runs <= 3  ~ "small",
    num_runs >= 4 & num_runs <= 20  ~ "medium",
    TRUE      ~ "large"  # Catches everything else
  )) %>%
  arrange(num_runs) %>%
  group_by(batches) %>%
  dplyr::select(Run, BioSample, Species) %>%
  group_walk(~ write_tsv(.x, paste0("../../config/samples_", .y$batches, ".tsv")))

runinfo_partitioned %>%
  group_by(batches) %>%
  summarize(n_distinct(Species))
```

# compare alignment diversity metrics to alignment-free diversity metrics

```{r}

align_div <- fread("../../data/current_most_complete_data.csv")

# Get a vector of all matching file paths
# file_list <- list.files(path = "../results/subset/kmer_distances", 
#                         pattern = ".txt", 
#                         full.names = TRUE)
file_list <- list.files(path = "../results/nofilt/subset/kmer_distances", 
                        pattern = ".txt", 
                        full.names = TRUE)

kmer_div <- file_list %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("bray_curtis", "cosine")), .id = "source_file") %>%
  mutate(phylo_name = gsub(".*kmer_distances/|.txt","",source_file))

all_div <- align_div %>%
  left_join(kmer_div, by = "phylo_name") %>%
  dplyr::select(phylo_name, Avg_Val_pi, Avg_Val_het, bray_curtis, cosine, final_area_km2, height_clean, genome_size, redlistCategory) %>%
  mutate(R_h2 = final_area_km2/((height_clean/1000)^2))

all_div$redlistCategory[(all_div$redlistCategory == "Lower Risk/near threatened")] <- "Near Threatened"
all_div$redlistCategory[(all_div$redlistCategory == "Lower Risk/least concern")] <- "Least Concern"

all_div$redlistCategory <- factor(all_div$redlistCategory, levels = c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered", "Extinct in the Wild"))

all_div$redlistCategoryCollapsed <- as.character(all_div$redlistCategory)
all_div$redlistCategoryCollapsed[(all_div$redlistCategoryCollapsed %in% c("Least Concern", "Near Threatened"))] <- "Non-Threatened"
all_div$redlistCategoryCollapsed[(all_div$redlistCategoryCollapsed %in% c("Vulnerable", "Endangered", "Critically Endangered", "Extinct in the Wild"))] <- "Threatened"
all_div$redlistCategoryCollapsed <- factor(all_div$redlistCategoryCollapsed, levels = c("Non-Threatened", "Threatened"))

# initial data analysis
ggplot(all_div, aes(x = Avg_Val_pi, y = Avg_Val_het)) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_cor()

ggplot(all_div, aes(x = Avg_Val_pi, y = bray_curtis)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple") +
  stat_cor() +
  labs(x = "Mean nucleotide diversity", y = "Mean Bray-Curtis\nk-mer dissimilarity")

ggsave(paste0("../../results/by-date/", today, "/nucleotide_diversity_vs_bray_curtis.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = Avg_Val_het, y = bray_curtis)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple") +
  stat_cor() +
  labs(x = "Mean heterozygosity", y = "Mean Bray-Curtis\nk-mer dissimilarity")

ggsave(paste0("../../results/by-date/", today, "/heterozygosity_vs_bray_curtis.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = Avg_Val_pi, y = cosine)) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_cor()

ggplot(all_div, aes(x = Avg_Val_het, y = cosine)) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_cor()

# correlations with threat status
ggplot(all_div, aes(x = redlistCategory, y = Avg_Val_het)) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggforce::geom_sina(
    maxwidth = .6, scale = "count", seed = 1,
    size = 2, alpha = .5
  ) +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Red List category", y = "Heterozygosity")

ggsave(paste0("../../results/by-date/", today, "/redlist_vs_heterozygosity.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = redlistCategory, y = Avg_Val_pi)) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggforce::geom_sina(
    maxwidth = .6, scale = "count", seed = 1,
    size = 2, alpha = .5
  )+
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Red List category", y = "Nucleotide diversity")

ggsave(paste0("../../results/by-date/", today, "/redlist_vs_pi.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = redlistCategory, y = bray_curtis)) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggforce::geom_sina(
    maxwidth = .6, scale = "count", seed = 1,
    size = 2, alpha = .5
  ) +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Red List category", y = "Bray-Curtis\nk-mer dissimilarity") 

ggsave(paste0("../../results/by-date/", today, "/redlist_vs_bray_curtis.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = redlistCategoryCollapsed, y = Avg_Val_het)) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggforce::geom_sina(
    maxwidth = .6, scale = "count", seed = 1,
    size = 2, alpha = .5
  ) +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Red List category", y = "Heterozygosity") 

ggsave(paste0("../../results/by-date/", today, "/redlist_collapsed_vs_heterozygosity.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = redlistCategoryCollapsed, y = Avg_Val_pi)) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggforce::geom_sina(
    maxwidth = .6, scale = "count", seed = 1,
    size = 2, alpha = .5
  ) +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Red List category", y = "Nucleotide diversity") 

ggsave(paste0("../../results/by-date/", today, "/redlist_collapsed_vs_pi.pdf"), width = 6, height = 5)

ggplot(all_div, aes(x = redlistCategoryCollapsed, y = bray_curtis)) +
  geom_violin(draw_quantiles = c(0.5)) +
  ggforce::geom_sina(
    maxwidth = .6, scale = "count", seed = 1,
    size = 2, alpha = .5
  ) +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Red List category", y = "Bray-Curtis\nk-mer dissimilarity") 

ggsave(paste0("../../results/by-date/", today, "/redlist_collapsed_vs_bray_curtis.pdf"), width = 6, height = 5)

# correlations with range size
ggplot(all_div, aes(x = Avg_Val_pi, y = R_h2)) +
  geom_point() +
  scale_y_log10() +
  geom_smooth(method = "lm") +
  stat_cor()

ggplot(all_div, aes(x = Avg_Val_het, y = R_h2)) +
  geom_point() +
  scale_y_log10() +
  geom_smooth(method = "lm") +
  stat_cor()

ggplot(all_div, aes(x = bray_curtis, y = R_h2)) +
  geom_point() +
  scale_y_log10() +
  geom_smooth(method = "lm") +
  stat_cor()

ggplot(all_div, aes(x = cosine, y = R_h2)) +
  geom_point() +
  scale_y_log10() +
  geom_smooth(method = "lm") +
  stat_cor()

```

# Phylogenetic analysis
```{r}
jtree <- read.tree("../../resources/analysis_tree.tre")

plot(jtree)
jtree$node.label <- NULL

all_div <- all_div %>%
  filter(phylo_name %in% jtree$tip.label) %>%
  arrange(match(phylo_name, jtree$tip.label)) %>%
  mutate(redlistCategoryCollapsedFactor = factor(redlistCategoryCollapsed, levels = c("Non-Threatened", "Threatened")),
         redlistCategoryFactor = factor(redlistCategory, levels = c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered", "Extinct in the Wild")))

# Combine tree and data into a single comparative object
# data MUST be dataframe, NOT a tibble
comp_data <- comparative.data(phy = jtree, data = as.data.frame(all_div), 'phylo_name', na.omit = F)

# threat status
pgls_mod_bc <- pgls(bray_curtis ~ redlistCategoryCollapsedFactor, data = comp_data, lambda = "ML")
pgls_mod_pi <- pgls(log10(Avg_Val_pi) ~ redlistCategoryCollapsedFactor, data = comp_data, lambda = "ML")
pgls_mod_het <- pgls(log10(Avg_Val_het) ~ redlistCategoryCollapsedFactor, data = comp_data, lambda = "ML")

summary(pgls_mod_bc)
summary(pgls_mod_pi)
summary(pgls_mod_het)

brunch(bray_curtis ~ redlistCategoryFactor, data = comp_data)
brunch(Avg_Val_pi ~ redlistCategoryFactor, data = comp_data)

# Run the PGLS model
# Setting lambda = "ML" estimates Pagel's Lambda via Maximum Likelihood
pgls_mod_bray_curtis <- pgls(log10(R_h2) ~ scale(bray_curtis) + log10(genome_size), data = comp_data, lambda = "ML")
pgls_mod_pi <- pgls(log10(R_h2) ~ scale(log10(Avg_Val_pi)) + log10(genome_size), data = comp_data, lambda = "ML")

# View results
summary(pgls_mod_pi)
summary(pgls_mod_bray_curtis)

# 
all_div %>%
  filter(!is.na(Avg_Val_pi) & !is.na(R_h2)) %>%
  nrow()

all_div %>%
  filter(!is.na(bray_curtis) & !is.na(redlistCategoryCollapsed)) %>%
  count(redlistCategoryCollapsed)


all_div %>%
  filter(!is.na(Avg_Val_pi) & !is.na(final_area_km2)) %>%
  nrow()

all_div %>%
  filter(!is.na(bray_curtis) & !is.na(final_area_km2)) %>%
  nrow()

all_div %>%
  filter(!is.na(Avg_Val_pi) & !is.na(height_clean)) %>%
  nrow()

all_div %>%
  filter(!is.na(bray_curtis) & !is.na(height_clean)) %>%
  nrow()



# try just area
pgls_mod_bray_curtis_range <- pgls(log10(final_area_km2) ~ scale(bray_curtis)  + log10(genome_size), data = comp_data, lambda = "ML")
pgls_mod_pi_range <- pgls(log10(final_area_km2) ~ scale(log10(Avg_Val_pi)) + log10(genome_size), data = comp_data, lambda = "ML")

summary(pgls_mod_pi_range)
summary(pgls_mod_bray_curtis_range)

# try just height
pgls_mod_bray_curtis_height <- pgls(log10(height_clean) ~ scale(bray_curtis)  + log10(genome_size), data = comp_data, lambda = "ML")
pgls_mod_pi_height <- pgls(log10(height_clean) ~ scale(log10(Avg_Val_pi)) + log10(genome_size), data = comp_data, lambda = "ML")

summary(pgls_mod_pi_height)
summary(pgls_mod_bray_curtis_height)
```

# plot old k-mer distance script vs new script
```{r}

old_kmer_distances <- list.files(path = "../results/nofilt/subset/kmer_distances", 
                        pattern = ".*dist.txt", 
                        full.names = TRUE)

old_kmer_distances <- old_kmer_distances %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("bray_curtis", "cosine")), .id = "source_file") %>%
  mutate(phylo_name = gsub(".*kmer_distances/|_dist.txt","",source_file))

new_kmer_distances <- list.files(path = "../results/nofilt/kmer_distances", 
                        pattern = ".*dist.txt", 
                        full.names = TRUE)

new_kmer_distances <- new_kmer_distances %>% 
  set_names() %>% 
  map_dfr(~fread(.x, col.names = c("bray_curtis", "cosine")), .id = "source_file") %>%
  mutate(phylo_name = gsub(".*kmer_distances/|_dist.txt","",source_file))

new_vs_old_dist <- new_kmer_distances %>% 
  left_join(old_kmer_distances, by = "phylo_name")

ggplot(new_vs_old_dist, aes(x = bray_curtis.y, y = bray_curtis.x)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  labs(x = "Bray-Curtis distance (slow method)", y = "Bray-Curtis distance (fast method)", title = paste0("Validate better distance calculation\nn = ", sum(complete.cases(new_vs_old_dist))))

ggsave(paste0("../../results/by-date/", today, "/new_vs_old_bray_curtis.pdf"), width = 5, height = 5)
```

# k-mer mar
```{r}

kmer_tables_full <- list.files(path = "../results/nofilt/paste", pattern = ".txt.gz", full.names = T)

kmer_tables_full <- names(sort(sapply(kmer_tables_full, file.size)))

# file_path <- "../results/nofilt/paste/Oemleria_cerasiformis.txt.gz"
chunk_size <- 100000  # rows per chunk, tune to your memory

# function to iterate through chunks
sum_chunk <- function(chunk, pos) {
  
  cat("Pos: ", pos, "\n")
  chunk_no_kmers <- chunk[, 2:ncol(chunk)]
  
  chunk_sums <- colSums(chunk_no_kmers, na.rm = TRUE)
  chunk_num_non_zeros <- colSums(chunk_no_kmers != 0, na.rm = TRUE)
  
  if (is.null(running_sums)) {
    running_sums <<- chunk_sums
  } else {
    running_sums <<- running_sums + chunk_sums
  }
  
  if(is.null(running_num_non_zeros)){
    running_num_non_zeros <<- chunk_num_non_zeros
  } else {
    running_num_non_zeros <<- running_num_non_zeros + chunk_num_non_zeros
  }
}

# loop over k-mer tables and summarize coverage vs number of unique k-mers found
for(file_path in kmer_tables_full) {
  cat(file_path, "\n")
  
  # Accumulator for running column sums
  running_sums <- NULL
  running_num_non_zeros <- NULL
  
  read_tsv_chunked(
    file_path,
    callback = SideEffectChunkCallback$new(sum_chunk),
    chunk_size = chunk_size,
    col_types = cols(),  # let it guess types; or specify explicitly for speed/safety
    col_names = F
  )
  
  kmer_cov <- data.frame(
    file = file_path,
    sample = 1:length(running_sums),
    kmer_sum = running_sums,
    kmer_num_unique = running_num_non_zeros
  )
  
  write.table(
    kmer_cov,
    paste0(
      "../../results/by-date/",
      today,
      "/kmer_coverage_summary.txt"
    ),
    append = T,
    quote = F,
    row.names = F
  )
}

ggplot(kmer_cov, aes(x = kmer_sum, y = kmer_num_unique)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()

print(running_sums)
print(running_num_non_zeros)
```

# k-mer mar (mutations-area relationship)

Input: the subsetted k-mer tables from the `kmers_nofilt_only` branch
(`results/nofilt/subset/kmer_tables/{species}.txt`). These are space-delimited
matrices with one row per sampled k-mer, one column per sample, k-mer names
already stripped (see `workflow/rules/kmc_nofilt.smk:120`).

Encoding: counts are binarized to 0/1 for k-mer absence/presence per sample,
which is a valid ploidy-1 allele-count matrix for `mar`'s `text_parser()`
(values in {0, ploidy}).

Caveat: `mar`'s theory assumes rows are independent biallelic SNP sites. K-mers
violate this (one SNP generates k overlapping k-mers), so the fitted z is a
heuristic "k-mer-area" exponent, not a true MAR exponent. Coordinates are not
required: we fit the theoretical MAR (`MARtheory()`) against sample size
(`Atype = "N"`), which needs no map.

## install mar (one-time; run interactively, not knit)

```{r, eval = FALSE}
# install mar and its dependencies (README: github.com/meixilin/mar).
# terra pulls in gdal; if conda is preferred instead:
#   mamba create --name mar r-base r-terra r-matrixstats r-sars gdal
# then remotes::install_github("meixilin/mar") inside that env.
install.packages(c("sars", "matrixStats", "terra"))
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("meixilin/mar")
```

## load mar and choose species

```{r}
if (!requireNamespace("mar", quietly = TRUE)) {
  stop("Package 'mar' is not installed. Run the install chunk above first.")
}
suppressPackageStartupMessages(library(mar))

# directory holding the subsetted k-mer tables (paths relative to workflow/)
mar_kmer_dir <- "../results/nofilt/subset/kmer_tables"

# all species with a subset table available
mar_species <- list.files(mar_kmer_dir, pattern = "\\.txt$", full.names = FALSE) %>%
  tools::file_path_sans_ext() %>%
  sort()

# species to analyze in the single-species chunks below
mar_species_sel <- mar_species[1]
cat("Available species:", paste(mar_species, collapse = ", "), "\n")
cat("Selected for single-species run:", mar_species_sel, "\n")

# results go under results/by-date/{today}/ (same convention as other chunks)
mar_out_dir <- file.path("../../results/by-date", as.character(today), "mar")
dir.create(mar_out_dir, recursive = TRUE, showWarnings = FALSE)
```

## binarize count table and format for mar

```{r}
# function: read a space-delimited subset k-mer table (no header, no k-mer
# column), binarize to 0/1 presence/absence, write a tab-delimited copy for
# text_parser() (its delimiter guesser only accepts tab or comma), and return
# the output path
mar_prep_input <- function(species, kmer_dir = mar_kmer_dir, out_dir = mar_out_dir) {
  in_path <- file.path(kmer_dir, paste0(species, ".txt"))
  stopifnot("subset k-mer table not found" = file.exists(in_path))

  # read full matrix; fread handles large files and spaces as separator
  geno_mat <- fread(in_path, sep = " ", header = FALSE) %>%
    as.matrix()

  # binarize: any nonzero count -> present (1); ploidy = 1
  geno_mat[geno_mat > 0] <- 1
  storage.mode(geno_mat) <- "integer"

  # mar's text_parser reads txt/csv/tsv with tab or comma delimiters and no
  # header/row names -- write a tab-delimited copy
  out_path <- file.path(out_dir, paste0("mar_input_", species, ".txt"))
  write.table(
    geno_mat,
    file = out_path,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
  cat("prepped:", out_path, "|", nrow(geno_mat), "k-mers x", ncol(geno_mat), "samples\n")
  out_path
}
```

## import genotypes into mar

```{r}
# text_parser() reads the binarized matrix as alternative-allele counts with
# ploidy = 1: 0 = k-mer absent, 1 = k-mer present. Sample IDs are optional; if
# omitted, samples are numbered 1..N. Chromosome/position files are omitted --
# they only annotate variant.id, which the theoretical workflow does not use.
mar_geno <- text_parser(
  geno.fn = mar_prep_input(mar_species_sel),
  ploidy = 1
)
print(mar_geno)
```

## build the genomaps object (technicality, no spatial meaning)

```{r}
# MARtheory() requires a genomaps object whose maps$sample.id length matches
# the genotype columns. The empirical sampling functions (MARsampling,
# MARextinction) are not used here, so the coordinates are dummies placed on an
# arbitrary plane (incrs = "", mapcrs = "") with no projection applied.
mar_dummy_coords <- data.frame(
  ID = paste0("s", seq_len(ncol(mar_geno$genotype))),
  LON = seq_len(ncol(mar_geno$genotype)),
  LAT = seq_len(ncol(mar_geno$genotype))
)

mar_maps <- marmaps(lonlatdf = mar_dummy_coords, mapcrs = "", incrs = "")

mar_gm <- genomaps(mar_geno, mar_maps)
print(mar_gm)
```

## site-frequency spectrum: observed vs neutral expectation

```{r}
# The SFS here is a histogram of k-mer presence counts: how many k-mers are
# present in exactly k of the N samples (k = 0..N at ploidy 1).
mar_obs_sfs <- sfs(mar_gm)
mar_exp_sfs <- expsfs(mar_gm)

plot(mar_obs_sfs, expected = mar_exp_sfs, log = "x",
     main = paste0("K-mer frequency spectrum: ", mar_species_sel))

# folded spectrum, useful when "ancestral state" (which sample is the reference)
# is arbitrary -- as it is here
plot(sfs(mar_gm, folded = TRUE), expected = expsfs(mar_gm, folded = TRUE), log = "x")
```

## theoretical mutations(k-mers)-area relationship

```{r}
# MARtheory() predicts the expected number of "mutations" (k-mers present in at
# least one sampled individual) as a function of subsample size n, given the
# full SFS. Assumptions: random mating, random spatial distribution,
# unlinked sites -- none strictly hold for k-mers, hence heuristic.
mar_theory_df <- MARtheory(mar_gm)
head(mar_theory_df)

plot(mar_theory_df, fit = TRUE,
     main = paste0("Theoretical k-mer MAR: ", mar_species_sel))
```

## fit the power-law model M = cA^z against sample size

```{r}
# Atype = "N" fits diversity against the number of sampled individuals -- this
# is what makes the coordinate-free workflow possible. Mtype = "M" is the
# number of mutations (unique k-mers captured).
mar_fit <- MARcalc(mar_theory_df, Mtype = "M", Atype = "N")
summary(mar_fit)

plot(mar_theory_df, fit = mar_fit$par, Mtype = "M", Atype = "N",
     main = paste0("K-mer MAR fit: ", mar_species_sel))

# export fit summary (c, z, p-values, R2)
mar_fit_out <- capture.output(summary(mar_fit))
writeLines(mar_fit_out,
           file.path(mar_out_dir, paste0("mar_fit_", mar_species_sel, ".txt")))
```

## loop: theoretical k-mer MAR for all species

```{r}
# Repeat the binarize -> import -> theory -> fit workflow for every species with
# a subset table, collecting a species-level summary of the scaling exponent z.
# Requires >= 4 samples per species for a meaningful z fit; species below that
# threshold are skipped with a warning.
mar_run_species <- function(species, kmer_dir = mar_kmer_dir, out_dir = mar_out_dir) {
  cat("=== ", species, " ===\n")
  geno_path <- tryCatch(
    mar_prep_input(species, kmer_dir = kmer_dir, out_dir = out_dir),
    error = function(e) {
      warning("skipping ", species, ": ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(geno_path)) return(NULL)

  geno <- tryCatch(text_parser(geno.fn = geno_path, ploidy = 1),
                   error = function(e) {
                     warning("text_parser failed for ", species, ": ", conditionMessage(e))
                     NULL
                   })
  if (is.null(geno)) return(NULL)

  coords <- data.frame(
    ID = paste0("s", seq_len(ncol(geno$genotype))),
    LON = seq_len(ncol(geno$genotype)),
    LAT = seq_len(ncol(geno$genotype))
  )
  maps <- marmaps(coords, mapcrs = "", incrs = "")
  gm <- genomaps(geno, maps)

  if (length(maps$sample.id) < 4) {
    warning("skipping ", species, ": only ", length(maps$sample.id),
            " samples; need >= 4 for a stable z fit")
    return(NULL)
  }

  theory_df <- MARtheory(gm)
  fit <- MARcalc(theory_df, Mtype = "M", Atype = "N")
  if (is.null(fit)) {
    warning("MARcalc returned NULL for ", species)
    return(NULL)
  }

  fit_sum <- summary(fit)
  writeLines(capture.output(fit_sum),
             file.path(out_dir, paste0("mar_fit_", species, ".txt")))

  data.frame(
    species = species,
    n_samples = length(maps$sample.id),
    n_kmers = length(gm$geno$variant.id),
    c = fit_sum$Parameters[[1, "Estimate"]],
    z = fit_sum$Parameters[[2, "Estimate"]],
    z_p = fit_sum$Parameters[[2, "Pr(>|t|)"]],
    R2_adj = fit_sum$R2a
  )
}

mar_z_table <- set_names(mar_species, mar_species) %>%
  map_dfr(mar_run_species)

# one row per species: the k-mer-area scaling exponent z for downstream
# comparisons (e.g. join to the threat-status / range-size analysis chunks)
mar_z_table

write.table(
  mar_z_table,
  file.path(mar_out_dir, "mar_z_summary.txt"),
  quote = FALSE,
  row.names = FALSE,
  sep = "\t"
)
```
