---
title: "R Notebook"
output: html_notebook
---

# set-up

```{r}
rm(list = ls())
gc()
.rs.restartR()
```

# packages

```{r}
library(tidyverse)
library(data.table)
```

# compile jules' lists

```{r}
# load 
pt1 <- fread("../../data/fastq_master_mapped_pt1.csv")
pt2 <- fread("../../data/fastq_species_master_mapped_part2_filtered.csv")
all_sam <- rbind(pt1, pt2)

# basic stats
species_with_many_samples <- all_sam %>%
  group_by(Species) %>%
  filter(n() > 1)

# list of species where runs collapsed to biosamples
collapsed_sam_species <- unique(all_sam[grepl("SAMN", all_sam$SRR), "Species"])

# runs for species without collapsed samples
non_collapsed_sam <- all_sam[!(all_sam$Species %in% collapsed_sam_species$Species),]
```

## output biosamples to be processed with entrez

```{r}
# output SAMN
write.table(all_sam[grepl("SAMN|SAME|SAMD", all_sam$SRR), "SRR"], "../../resources/biosample_ids.txt", quote = F, row.names = F, col.names = F)
```

## now grab run ids associated with biosamples

## get SRR associated with SAMN 

remove any SRR that are not matching jules' filters

awk -F, 'NR==1 || ($13 == "WGS" && $15 == "GENOMIC" && $16 == "PAIRED" && $19 == "ILLUMINA")' SraRunInfo.csv > filtered_SraRunInfo.csv

```{r}
runinfo <- fread("/global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources/runinfo.csv")

runinfo_filt <- runinfo %>% filter(LibraryStrategy == "WGS" & LibrarySource == "GENOMIC" & LibraryLayout == "PAIRED" & Platform == "ILLUMINA")

all_sam_run_only <- merge(all_sam, runinfo_filt[,c("Run", "BioSample")], by.x = "SRR", by.y = "BioSample", all.x = T)

# if run starts with SAMN, replace with real Run number
samn_bool <- grepl("SAMN|SAME|SAMD", all_sam_run_only$SRR)
all_sam_run_only$SRR[samn_bool] <- all_sam_run_only$Run[samn_bool]

# expect TRUE
if(!all(all_sam_run_only$SRR == all_sam_run_only$Run, na.rm = T)){
  stop("SRR does not match Run")
}

```

## now output the runs and grab all of their biosample ids

My hope is to standardize the info, so it's intuitive, rather than having my grouping variable sometimes be the run id and sometimes be the biosample id

```{r}
write.table(all_sam_run_only[,1], "../../resources/runs_srr_only.txt", quote = F, row.names = F, col.names = F)
```

## now grab all run info for every run

```{bash, engine.opts='-l'}

cd /global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources
pwd

ENV_BIN="/global/home/users/milesroberts/.local/share/mamba/envs/entrez/bin"

readarray -t runs < "runs_srr_only.txt"

# Clear out the old file if it exists so you don't keep appending duplicates
> runinfo_all_runs.csv

for run in "${runs[@]}"; do
    echo "$run"
    $ENV_BIN/esearch -db sra -query $run | $ENV_BIN/efetch -format runinfo >> runinfo_all_runs.csv
    sleep 1
done
```

## Finally, organize runinfo into one table with three columns: run, biosample, species

```{r}
runinfo <- read.csv("/global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources/runinfo_all_runs_unique.csv", fill = T, header = F)

# run_list <- read.csv("/global/scratch/users/milesroberts/moi_lab_projects/panNrefNkmer/resources/runs_srr_only.txt", fill = T, header = F)
# 
# sum(run_list$V1 %in% runinfo$V1)

runinfo_sorted <- runinfo %>% 
  #filter(!grepl("metagenome|uncultured|unidentified|human|synthetic|prokaryotic|mixed libraries|Mus musculus|Homo sapiens|Saccharomyces cerevisiae|Saccharomonospora azurea|Caenorhabditis elegans|Bos taurus|Gallus gallus|Monodelphis domestica|Drosophila", V29)) %>%
  #filter(V29 %in% (all_sam$Species)) %>% 
  filter(V13 == "WGS" & V15 == "GENOMIC" & V16 == "PAIRED" & V19 == "ILLUMINA") %>%
  dplyr::select(V1, V26, V29) %>% 
  arrange(V26, V1, V29) 

runinfo_sorted <- runinfo_sorted[-1:-5,]

# replace space with _
runinfo_sorted$V29 <- gsub(" ", "_", runinfo_sorted$V29)
names(runinfo_sorted) = c("Run", "BioSample", "Species")

# remove species with only one run
# cannot calculate k-mer distances for such species
# sort species by number of runs (analyze species with less data first)
runinfo_filtered <- runinfo_sorted %>%
  group_by(Species) %>%
  filter(n_distinct(BioSample) > 1) %>%
  ungroup() %>%
  droplevels() %>%
  filter(Species %in% (all_sam$Species)) %>% 
  arrange(Species)
  #group_by(Species) %>%
  #mutate(n_runs = n()) %>% 
  #arrange(n_runs, Species)%>%
  #dplyr::select(Run, BioSample, Species)

# summary statistics
cat("Number of Species:", n_distinct(runinfo_filtered$Species), "\n")
cat("Number of BioSamples:", n_distinct(runinfo_filtered$BioSample), "\n")
cat("Number of Runs:", n_distinct(runinfo_filtered$Run), "\n")

# check for species not in original Jules list
# seems like my list excludes some species that are cultivars
runinfo_filtered$Species[!(runinfo_filtered$Species %in% species_with_many_samples$Species)]

unique(species_with_many_samples$Species[!(species_with_many_samples$Species %in% runinfo_filtered$Species)])

# save list of species
write.table(runinfo_filtered, "../../config/samples_all.tsv", quote = F, row.names = F, col.names = T, sep = "\t")

# partition species into groups with a small, medium, and large amount of data
# will process species with very little data first

runinfo_partitioned <- runinfo_filtered %>%
  group_by(Species) %>%
  mutate(num_runs = n(),
         batches = case_when(
    num_runs <= 3  ~ "small",
    num_runs >= 4 & num_runs <= 20  ~ "medium",
    TRUE      ~ "large"  # Catches everything else
  )) %>%
  arrange(num_runs) %>%
  group_by(batches) %>%
  dplyr::select(Run, BioSample, Species) %>%
  group_walk(~ write_tsv(.x, paste0("../../config/samples_", .y$batches, ".tsv")))

runinfo_partitioned %>%
  group_by(batches) %>%
  summarize(n_distinct(Species))
```

